Improved state-space models for inference about spatial and temporal variation in abundance

### 2012-09-05 Dail_Mad_ext.tex
intro and lit

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}

@@

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in abundance, for
example to test for the existence of density-dependence population
growth \citep{dennis_taper:1994,viljugrein_etal:2005}. Such models are
also required in applied ecological research
for predicting species' responses to environmental change or
conservation actions. For example, %many researchers are currently
models of population dynamics allow for inference about the effects of
of climate change on population viability.
%investigations of the influence of climate on population dynamics so that
%predictions can be made about the effects of climate change.
Several complicating factors must be confronted when
addressing these questions. First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions. Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection. Failure to account for process variation and observation
error can bias the estimators of abundance and related parameters.

State-space models have become the most widespread approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_taper:1994,
  dennis_etal:2006}. Conventional state-space models are
simply time-series models in which the true state of the system is
observed imperfectly. Originally developed for xxx (cite Kalman filter
papers), the ability to model both the ecological state process and
the observation process has made them relevant in numerous ecological
applications [need to clean this up]. State-space models have been
defined differently by different authors, but we follow
\citet{buckland_etal:2004} the definition : ``xxx'' . Typically, the
ecological
process is described by a Markovian model that includes a
deterministic description of population change as well as two random
sources of variation for ``process variation''. The model for the
observation process is typically phenomenological in the sense that
observation error is not defined explicitly. As an example, let $N_t$ be
the abundance of a species during year $t$, and $X_t$ be the observed data,
which differs from  due to observation error ($\zeta$). A simple state
space model that includes exponential growth and process variation
($\grave{o}$) could be written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
X_t = N_t + \zeta_t
\end{gather}
Here   is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions: $\grave{o}_t \sim
\mathrm{N}(0, \sigma)$ and $\zeta_t \sim \mathrm{N}(0, \tau)$
\citep{devalpine_hastings:2002,dennis_etal:2006}. Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace $N_t$ with $\log(Y_t/A)$   where $A$ is the
area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.

>>>>

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}

@@

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance---hypotheses relating to the importance and existence of
phenomenon such as density-dependent population regulation, population
cycling, and spatial synchrony
\citep{royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999}.
%growth ,viljugrein_etal:2005}.
In applied contexts, population models are used for %population viability
%analysis, including the estimation of
estimating extinction probabilities
\citep{nadeem_lele:2011} and for predicting the effects of future
environmental conditions or conservation actions on population size
\citep{jamieson_brooks:2004}.
%Such models are
%also required in applied ecological research
%for predicting species' responses to environmental change or
%conservation actions. For example, %many researchers are currently
%models of population dynamics allow for inference about the effects of
%of climate change on population viability.
%investigations of the influence of climate on population dynamics so that
%predictions can be made about the effects of climate change.
In order to address these questions, several complicating factors must
be confronted when fitting population models to data.
First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{kery_etal:2009}. Failure to account for process
variation and observation error can bias the estimators of abundance
and related parameters.

State-space models are a widely used approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_taper:1994,
  dennis_etal:2006}. Conventional state-space models are
simply time-series models in which the true state of the system
(e.g. population size during each year) is
observed imperfectly. Originally developed for xxx (cite Kalman filter
papers), the ability to model both the ecological state process and
the observation process has made them relevant in numerous ecological
applications [need to clean this up]. State-space models have been
defined differently by different authors, but we follow
\citet{buckland_etal:2004} the definition : ``xxx'' . Typically, the
ecological
process is described by a Markovian model that includes a
deterministic description of population change as well as two random
sources of variation for ``process variation''. The model for the
observation process is typically phenomenological in the sense that
observation error is not defined explicitly. As an example, let $N_t$ be
the abundance of a species during year $t$, and $X_t$ be the observed data,
which differs from  due to observation error ($\zeta$). A simple state
space model that includes exponential growth and process variation
($\grave{o}$) could be written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
X_t = N_t + \zeta_t
\end{gather}
Here   is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions: $\grave{o}_t \sim
\mathrm{N}(0, \sigma)$ and $\zeta_t \sim \mathrm{N}(0, \tau)$
\citep{devalpine_hastings:2002,dennis_etal:2006}. Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace $N_t$ with $\log(Y_t/A)$   where $A$ is the
area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.
